---
title: "Optional: Eigen Decomposition"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: inline
---

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
xaringanExtra::use_panelset()

```

A correlation matrix is a square matrix (i.e. same number of columns as rows), and it is symmetric on the diagonals. Furthermore, the diagonal values themselves are all 1 (the correlation of a variable with itself is 1). 

So we might see a correlation matrix (lets call it $\mathbf{R}$) that looks like this:  

$$
\begin{equation*}
\mathbf{R} = 
\begin{bmatrix}
1.00, 0.72, 0.63, 0.54, 0.45, 0.36 \\
0.72, 1.00, 0.56, 0.48, 0.40, 0.32 \\
0.63, 0.56, 1.00, 0.42, 0.35, 0.28 \\
0.54, 0.48, 0.42, 1.00, 0.30, 0.24 \\
0.45, 0.40, 0.35, 0.30, 1.00, 0.20 \\
0.36, 0.32, 0.28, 0.24, 0.20, 1.00 \\
\end{bmatrix} 
\end{equation*} 
$$

One way in which we can create a symmetric square matrix is to take the product of some vector $\mathbf{f}$ with $\mathbf{f'}$ ($\mathbf{f}$ "transposed"). This is just how you might "square" a vector. To multiply two vectors you take each entry of the first and multiply it with each entry of the second.  

For instance:  
$$
\begin{equation*}
\mathbf{f} = 
\begin{bmatrix}
0.88 \\
0.83 \\
0.77 \\
0.69 \\
0.60 \\
0.50 \\
\end{bmatrix} 
\qquad 
\mathbf{f} \mathbf{f'} = 
\begin{bmatrix}
0.88 \\
0.83 \\
0.77 \\
0.69 \\
0.60 \\
0.50 \\
\end{bmatrix} 
\begin{bmatrix}
0.88, 0.83, 0.77, 0.69, 0.60, 0.50 \\
\end{bmatrix} 
\qquad = \qquad
\begin{bmatrix}
0.78, 0.74, 0.68, 0.61, 0.53, 0.44 \\
0.74, 0.69, 0.64, 0.58, 0.50, 0.42 \\
0.68, 0.64, 0.59, 0.53, 0.47, 0.39 \\
0.61, 0.58, 0.53, 0.48, 0.42, 0.35 \\
0.53, 0.50, 0.47, 0.42, 0.37, 0.30 \\
0.44, 0.42, 0.39, 0.35, 0.30, 0.25 \\
\end{bmatrix} 
\end{equation*}
$$


Note that I've chosen values for $\mathbf{f}$ that get pretty close to re-creating $\mathbf{R}$ (clever me!). I can get even closer if I take this and add to it the product of __another__ vector:  


$$
\begin{align}
\begin{bmatrix}
0.78, 0.74, 0.68, 0.61, 0.53, 0.44 \\
0.74, 0.69, 0.64, 0.58, 0.50, 0.42 \\
0.68, 0.64, 0.59, 0.53, 0.47, 0.39 \\
0.61, 0.58, 0.53, 0.48, 0.42, 0.35 \\
0.53, 0.50, 0.47, 0.42, 0.37, 0.30 \\
0.44, 0.42, 0.39, 0.35, 0.30, 0.25 \\
\end{bmatrix} 
&+
\begin{bmatrix}
0.06 \\
0.07 \\
0.09 \\
0.13 \\
0.27 \\
-0.85 \\
\end{bmatrix} 
\begin{bmatrix}
0.06, 0.07, 0.09, 0.13, 0.27, -0.85 \\
\end{bmatrix} 
\qquad = \qquad \\
\quad \\
\quad \\
\begin{bmatrix}
0.78, 0.74, 0.68, 0.61, 0.53, 0.44 \\
0.74, 0.69, 0.64, 0.58, 0.50, 0.42 \\
0.68, 0.64, 0.59, 0.53, 0.47, 0.39 \\
0.61, 0.58, 0.53, 0.48, 0.42, 0.35 \\
0.53, 0.50, 0.47, 0.42, 0.37, 0.30 \\
0.44, 0.42, 0.39, 0.35, 0.30, 0.25 \\
\end{bmatrix} 
&+
\begin{bmatrix}
0.00, 0.00, 0.00, 0.01, 0.02, 0.05 \\
0.00, 0.00, 0.01, 0.01, 0.02, 0.06 \\
0.00, 0.01, 0.01, 0.01, 0.02, 0.07 \\
0.01, 0.01, 0.01, 0.02, 0.03, 0.11 \\
0.02, 0.02, 0.02, 0.03, 0.07, 0.23 \\
0.05, 0.06, 0.07, 0.11, 0.23, 0.72 \\
\end{bmatrix} 
\qquad = \qquad
\begin{bmatrix}
0.78, 0.74, 0.68, 0.62, 0.55, 0.40 \\
0.74, 0.70, 0.65, 0.59, 0.52, 0.36 \\
0.68, 0.65, 0.60, 0.55, 0.49, 0.31 \\
0.62, 0.59, 0.55, 0.50, 0.45, 0.24 \\
0.55, 0.52, 0.49, 0.45, 0.44, 0.07 \\
0.40, 0.36, 0.31, 0.24, 0.07, 0.97 \\
\end{bmatrix} 
\end{align}
$$


I haven't just chosen a random additional vector here, I've chosen one that is "orthogonal" to the first vector (this gets a little complicated and trigonometry-related, but [click here](https://textbooks.math.gatech.edu/ila/dot-product.html){target="_blank"} for an explanation of why the dot product of orthogonal vectors is equal to zero).  


$$
\begin{align}
\begin{bmatrix}
0.88 \\
0.83 \\
0.77 \\
0.69 \\
0.60 \\
0.50 \\
\end{bmatrix} 
\cdot
\begin{bmatrix}
0.06 \\
0.07 \\
0.09 \\
0.13 \\
0.27 \\
-0.85 \\
\end{bmatrix} =
0
\end{align}
$$

In essence eigen decomposition is about finding a set of vectors that can, together, express our correlation matrix $\mathbf{R}$.  

For an $n \times n$ correlation matrix $\mathbf{R}$, there is a set of $n$ **eigenvectors** $x_i$ that solve the equation 
$$
\mathbf{x_i R} = \lambda_i \mathbf{x_i}
$$
Where the vector multiplied by the correlation matrix is equal to some **eigenvalue** $\lambda_i$ multiplied by that vector.  
We can write this without subscript $i$ as: 
$$
\begin{align}
& \mathbf{R X} = \mathbf{X \lambda} \\
& \text{where:} \\
& \mathbf{R} = \text{correlation matrix} \\
& \mathbf{X} = \text{matrix of eigenvectors} \\
& \mathbf{\lambda} = \text{vector of eigenvalues}
\end{align}
$$
these vectors that make up $\mathbf{X}$ are orthogonal [($\mathbf{XX' = I}$)](https://miro.medium.com/max/700/1*kyg5XbrY1AOB946IE5nWWg.png), which means that $\mathbf{R = X \lambda X'}$


For some demonstration in R, let's create a correlation matrix
```{r}
# lets create a correlation matrix, as f f'
f <- seq(.9,.4,-.1)
R <- f %*% t(f)
#give rownames and colnames
rownames(R)<-colnames(R)<-paste0("V",seq(1:6))
#constrain diagonals to equal 1
diag(R)<-1
# here's our correlation matrix!
R
```

The actual solving of the eigen decomposition is too much for me, but luckily the function `eigen()` does it for us!  
```{r}
# do eigen decomposition
e <- eigen(R)
print(e, digits=2)
```

It returns eigenvalues and corresponding eigenvectors.   
The eigenvectors are orthogonal ($\mathbf{XX' = I}$):
```{r}
round(e$vectors %*% t(e$vectors),2)
```

The eigenvectors represent the _direction_ of the vectors, and the eigenvalues represent *magnitude* (i.e., how much variance of $\mathbf{R}$ they capture).  

Our Principal Components $\mathbf{P}$ are the eigenvectors scaled by the square root of the eigenvalues:
```{r}
# eigenvectors
# e$vectors
# scaled by sqrt of eigenvalues
# diag(sqrt(e$values))

P <- e$vectors %*% diag(sqrt(e$values))
P
```

$$
R = PP'

\lambda X


$$

And with all of them, we can perfectly reproduce our correlation matrix: $\mathbf{R = PP'}$. 
```{r}
P %*% t(P)
```


But we don't want to keep all of them, because then we are not reducing anything. So we might choose to keep, e.g., just one.  
 
When "doing PCA" with the `principal()` function, we retain only one component with `nfactors = 1`:
```{r}
library(psych)
pc1<-principal(R, nfactors = 1, covar = FALSE, rotate = 'none')
pc1
```

Look familiar? It looks like the first component we computed manually. The first column of $\mathbf{P}$. The direction is flipped, but that is somewhat irrelevant because it's the same dimension (i.e., South is just "negative North"!)  
```{r}
cbind(pc1$loadings, P=P[,1])
```
We can now ask "how well does this component (on its own) recreate our correlation matrix?" 
```{r}
P[,1] %*% t(P[,1])
```
It looks close, but not quite. How much not quite? Measurably so!
```{r}
R - (P[,1] %*% t(P[,1]))
```

Notice the values on the diagonals of $\mathbf{p_1}\mathbf{p_1}'$.
```{r}
diag(P[,1] %*% t(P[,1]))
```
These aren't 1, like they are in $R$.  
But they are proportional: this is the amount of variance in each observed variable that is explained by this first component.  
The `principal()` function calls this 'communalities', which is really more of a term we see in factor analysis.  
```{r}
pc1$communality
```
1 minus these values is the unexplained variance, or 'uniqueness':  
```{r}
1 - diag(P[,1] %*% t(P[,1]))
pc1$uniquenesses
```
