---
title: "MSMR"
subtitle: "2022/23"
date: ""
output:
  rmdformats::material:
    highlight: kate
    self_contained: true
    # code_folding: show
    thumbnails: true
    gallery: true
    fig_width: 4
    fig_height: 4
    df_print: kable
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{css}
.header-panel {
    background-color: #a41ae4;
}

a {
  color: #a41ae4;
}
```


# Course materials

_Please note: materials are added incrementally as this course progresses through the year._

```{r echo=FALSE}
library(tidyverse)
library(kableExtra)

course <- tribble(~a, ~b, ~c, ~d)

course <- course %>%
    bind_rows(
        c(a = 'Week 1',
          b = 'Introduction to MLM',
          c = '[Slides](https://uoepsy.github.io/msmr/2223/lectures/msmr_lec01_IntroMLM.html#1)',
          d = '[Lab](https://uoepsy.github.io/msmr/2223/labs/01_intromlm.html)')
    ) %>%
    bind_rows(
        c(a = 'Week 2',
          b = 'Longitudinal and Logistic MLM',
          c = '[Slides A](https://uoepsy.github.io/msmr/2223/lectures/msmr_lec02_LogisticMLM.html#1)
          <br>
          [Slides B](https://uoepsy.github.io/msmr/2223/lectures/msmr_lec02-2_LinearLDA.html#1)',
          d = '[Lab](https://uoepsy.github.io/msmr/2223/labs/02_lmm_log.html)')
    ) %>%
    bind_rows(
        c(a = 'Week 3',
          b = 'Nonlinear longitudinal MLM',
          c = '[Slides](https://uoepsy.github.io/msmr/2223/lectures/msmr_lec03_NonLinearLDA.html#1)',
          d = '[Lab](https://uoepsy.github.io/msmr/2223/labs/03_nonlin.html)')
    ) %>%
    bind_rows(
        c(a = 'Week 4',
          b = 'Other Random Effects',
          c = '[Slides A](https://uoepsy.github.io/msmr/2223/lectures/msmr_lec04-1_RanEf_3lev.html#1)
          <br>
          [Slides B](https://uoepsy.github.io/msmr/2223/lectures/msmr_lec04-2_RanEf_Crossed.html#1)',
          d = '[Lab](https://uoepsy.github.io/msmr/2223/labs/04_ranef.html)')
    ) %>%
    bind_rows(
        c(a = 'Week 5',
          b = 'Recap & Individual Differences',
          c = '[Slides A](https://uoepsy.github.io/msmr/2223/lectures/msmr_lec05-1_IndivDiffs_Ext.html#1)
          <br>
          [Slides B](https://uoepsy.github.io/msmr/2223/lectures/msmr_lec05-2_IndivDiffs_Int.html#1)',
          d = '[Lab](https://uoepsy.github.io/msmr/2223/labs/05_multilevel_recap.html)')
    )

names(course) <- c('Week', 'Topic', 'Lecture slides', 'Lab materials')

course %>%
    kable(format="html", escape=FALSE) %>%
    kable_styling(full_width = FALSE)
```



# Course description

Multivariate Statistics and Methodology using R (MSMR) is a semester long course providing an advanced level overview of statistical analysis techniques and methodology issues relevant to psychological research. The course builds on the concepts and skills developed in Univariate Statistics and Methodology using R.



# Archive

- [Year 2021/22](https://uoepsy.github.io/msmr/2122/index.html)
- [Year 2020/21](https://uoepsy.github.io/msmr/2021/index.html)


# Site map

[Home](https://uoepsy.github.io/) > [MSMR](https://uoepsy.github.io/msmr/)
