```{r, echo=FALSE,message=FALSE,warning=FALSE}
HIDDEN_SOLS=FALSE
library(tidyverse)
```

```{r}
knitr::opts_chunk$set(fig.align = 'center')
```


# SEM II

### Packages {-}  

```{r message=FALSE}
library(psych)
library(lavaan)
library(semPlot)
```

### Lecture Slides {-}  

+ The lecture slides can be accessed [here](https://uoe-psychology.github.io/uoe_psystats/multivar/lectures/week11_sem2.pdf).   


## Overview {-}
<div class="noteBox">

In this final week we are going to address some practical issues that commonly arise in Structural Equation Modeling: Non-normality, ordinal-categorical endogeneous variables, and missingness. In the example for this week a researcher is interested in the effects of prenatal stress on child cognitive outcomes. She has a 5-item measure of prenatal stress and a 5 subtest measure of child cognitive ability, collected for 500 mother-infant dyads. 

+ The data is available as a .csv file here: [https://edin.ac/3bod7nJ](https://edin.ac/3bod7nJ)

</div>

## Exercises {-}

`r msmbstyle::question_begin(header="&#x25BA; Question 1")`
The prenatal stress questionnaire items are measured on a 3-point scale.  
  
Read in the data and fit a one-factor confirmatory factor analysis using an appropriate estimation method.
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
```{r message=FALSE}
stress_IQ_data <- read_csv("https://edin.ac/3bod7nJ")

head(stress_IQ_data)

describe(stress_IQ_data)
```

```{r categorical estimation}
# specify the model
model_stress <- 'Stress=~stress1+stress2+stress3+stress4+stress5'

# estimate the model - cfa will automatically switch to a categorical estimator 
# if we mention that our five variables are ordered-categorical, using the 'ordered' function
model_stress.est <- cfa(model_stress, data=stress_IQ_data,
                        ordered=c('stress1','stress2','stress3','stress4','stress5'))

# inspect the output
summary(model_stress.est, fit.measures=T, standardized=T)
```

We look at the 'robust' values for the fit statistics (those that appear in the right-hand column under 'robust') to check that our model fits well. Another new thing is the presence of 'thresholds' in the output. These are two thresholds per item in this example because we have a three-point response scale. Thresholds are the points on the hypothetical underlying continuum where individuals move from scoring in one category to the next one up. 
`r msmbstyle::solution_end()`
---
`r msmbstyle::question_begin(header="&#x25BA; Question 2")`
Now let's conduct a CFA for the IQ items. Check their distributions and fit a one-factor CFA using an appropriate estimation method
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`

We should check the item distributions for evidence of non-normality (skewness and kurtosis). We can use the describe() function from psych and plot the data using histograms or density curves.

```{r CFA for IQ items}
describe(stress_IQ_data[,c('IQ1','IQ2','IQ3','IQ4','IQ5')])

## GGPLOT
# temporarily reshape the data to long format to make it quicker to plot
stress_IQ_data %>% 
  pivot_longer(IQ1:IQ5, names_to="variable",values_to="score") %>%
  ggplot(aes(x=score))+
  geom_density()+
  facet_wrap(~variable)+
  theme_light()
```

Or we could do similar in base R: 
```{r eval=TRUE,echo=TRUE}
# specify that the figure is made up of 2-by-3, i.e. 6, panels, and the figure margins
par(mfrow = c(2,3), mar = c(5, 4, 1.5, 1.5))

## BASE R 
d_IQ1 <- density(stress_IQ_data$IQ1) # returns the density data
plot(d_IQ1, main='IQ1', xlab='score', frame=FALSE) # plots the results
d_IQ2 <- density(stress_IQ_data$IQ2)
plot(d_IQ2, main='IQ2', xlab='score', frame=FALSE)
d_IQ3 <- density(stress_IQ_data$IQ3)
plot(d_IQ3, main='IQ3', xlab='score', frame=FALSE)
d_IQ4 <- density(stress_IQ_data$IQ4)
plot(d_IQ4, main='IQ4', xlab='score', frame=FALSE)
d_IQ5 <- density(stress_IQ_data$IQ5)
plot(d_IQ5, main='IQ5', xlab='score', frame=FALSE)

# return to figures with a single panel
par(mfrow = c(1,1))
```

Because our variables seem to be non-normal, therefore, we should use a robust estimator such as MLR for our CFA

```{r robust estimator}
model_IQ <- 'IQ=~IQ1+IQ2+IQ3+IQ4+IQ5'

model_IQ.est <- cfa(model_IQ, data=stress_IQ_data, estimator='MLR')

summary(model_IQ.est, fit.measures=T, standardized=T)
```

The model doesn't fit very well so we could check the modification indices for local mis-specifications

```{r check mods}
modindices(model_IQ.est, sort=T)
```

It looks like we might need to include residual covariances between subtests 1 and 2 and between subtests 4 and 5, though we would want to double check this makes substantive sense (for example, do subtests 1 and 2 both measure memory while subtests 4 and 5 both test spatial ability?)


```{r make modifications}
model2_IQ <- 'IQ=~IQ1+IQ2+IQ3+IQ4+IQ5
IQ1~~IQ2
IQ4~~IQ5'
model2_IQ.est <- cfa(model2_IQ, data=stress_IQ_data, estimator='MLR')
summary(model2_IQ.est, fit.measures=T, standardized=T)
```

The fit of the model is now much improved!
`r msmbstyle::solution_end()`
---
`r msmbstyle::question_begin(header="&#x25BA; Question 3")`
Estimate the effect of prenatal stress on IQ using a full SEM
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
```{r full SEM}
SEM_model <- '
#IQ measurement model
IQ=~IQ1+IQ2+IQ3+IQ4+IQ5 
IQ1~~IQ2
IQ4~~IQ5

#stress measurement model 
Stress=~stress1+stress2+stress3+stress4+stress5 

#structural part of model
IQ~Stress'
```

We know that IQ is non-normal, so we would like to use a robust estimator (e.g. MLR). However, as lavaan will tell you if you try using `estimator="MLR"`, this is not yet supported for ordered data (i.e., the Stress items). It suggests instead using the WLSMV (weighted least square mean and variance adjusted) estimator:  
```{r}
SEM_model.est <- sem(SEM_model, data=stress_IQ_data,
                     ordered=c('stress1','stress2','stress3','stress4','stress5'),
                     estimator="WLSMV")

summary(SEM_model.est, fit.measures=T, standardized=T)
```

When we have *any* ordered-categorical endogenous variables in the model lavaan uses DWLS estimation, even if some of the variables are continous. We can see this from the 'Estimator' section of the output. We  can also see that the effect of prenatal stress on offspring IQ is $\beta = 0.388$ and statistically significant at $p<.05$.
`r msmbstyle::solution_end()`
---
`r msmbstyle::question_begin(header="&#x25BA; Question 4")`
In order to try and replicate the IQ CFA, our researcher collects a new sample of size $n=500$. However, she has some missing data. Specifically, those who scored poorly on earlier tests tended to feel discouraged and chose not to complete further tests.  
  
Conduct a CFA using the new data, taking account of the missingness.  
  
+ The data can be found at [https://edin.ac/2WGJav3](https://edin.ac/2WGJav3), and is in .csv format. 
 
`r msmbstyle::question_end()` 
`r msmbstyle::solution_begin(hidden=HIDDEN_SOLS)`
We can fit the model setting missing='FIML'. If data are missing at random (MAR) - i.e., missingness is related to the measured variables but not the unobserved missing values - then this gives us unbiased parameter estimates.  Unfortunately we can never know whether data are MAR for sure as this would require knowledge of the missing values. 

```{r message=FALSE}
IQ_data_new <- read_csv("https://edin.ac/2WGJav3")
```

```{r missingness}
IQ_model_missing <- 'IQ=~IQ1+IQ2+IQ3+IQ4+IQ5
IQ1~~IQ2
IQ4~~IQ5'

IQ_model_missing.est <- cfa(IQ_model_missing, data=IQ_data_new, missing='FIML', estimator="MLR")

summary(IQ_model_missing.est, fit.measures=T, standardized=T)
```
`r msmbstyle::solution_end()`
---
