---
title: "Week 5 Exercises: Assumptions, Diagnostics, Writing up"
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
library(ggdist)
xaringanExtra::use_panelset()
qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
library(lme4)
```

# Differences in Chimp Cognition  

:::frame
__Dataset: chimpcogdiff.csv__  

```{r}
ccd <- read_csv("../../data/chimpcogdiff.csv")
ccdape <- ccd |> count(zoo,apID)
```

While previous research has suggested that Chimpanzee cognitive ability is relatively stable across adulthood, less is known about the characteristics of chimpanzees that are associated with better cognitive performance and problem solving skills. The present research is interested in ascertaining whether a chimpanzee's level of dominance is associated with differences in cognitive ability, and whether this depends on the size of the enclosure in which a chimpanzee lives (i.e. whether dominance is associated with cognition to different extent depending on whether they live with a small number of other Chimpanzees, or live with a larger number).  

These data are from a study of `r nrow(ccdape)` adult chimpanzees studied across `r length(unique(ccdape$zoo))` research centers at zoos in the UK and Europe. Each chimpanzee was assessed on the "PCOG" - a measure of primate cognitive ability - on 3 occasions. Many of the participating chimpanzees took part in various research projects during their adolescence, and this has likely led to differences in both dominance and performance on these types of tasks.  

```{r}
#| echo: false
tibble(
  variable=names(ccd),
  description=c(
    "Zoo Name",
    "Zoo Location (UK or EU)",
    "Whether or not the chimpanzee took part in research during adolescence",
    "Size of enclosure in which the chimpanzee lives (enclosures tend to vary in the exact number of chimps living there, so these have been categorised as either 'small' or 'big' enclosures)",
    "Name of the chimpanzee",
    "Age (in years) of the chimpanzee",
    "Dominance Trait (Z-scored)",
    "Score on PCOG (scores range 0-35)"
  )
) |> gt::gt()
```


:::

`r qbegin(qcounter())`
Conduct an analysis to address the research aims!  


::: {.callout-tip collapse="true"}
#### Hints

- We have 3 observations per chimpanzee, and multiple chimps for each zoo.  
- Given the study background, why might it be a good idea to control for whether or not the chimps have taken part in research during adolescence? 

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
m1 = lmer(PCOG ~ zoo_loc + age + prev_ppt + DOM*enclosure_size + 
            (1 + DOM  | zoo) +
            (1 | zoo:apID), data = ccd)
summary(m1)
```


`r solend()`

`r qbegin(qcounter())`
Check the assumptions of your model




`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
plot(m1)
plot(m1,
     form = sqrt(abs(resid(.))) ~ fitted(.),
     type = c("p","smooth"))
qqnorm(ranef(m1)$`zoo:apID`[,1]);qqline(ranef(m1)$`zoo:apID`[,1])
qqnorm(ranef(m1)$zoo[,1]);qqline(ranef(m1)$zoo[,1])
qqnorm(ranef(m1)$zoo[,2]);qqline(ranef(m1)$zoo[,2])

library(performance)
check_predictions(m1)
```

`r solend()`

`r qbegin(qcounter())`
Check the extent to which your results may be sensitive to certain influential Xs or XXs. 
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
library(HLMdiag)
inf1 <- hlm_influence(m1,level=1)
dotplot_diag(inf1$cooksd, cutoff="internal")
inf2 <- hlm_influence(m1,level="zoo:apID")
dotplot_diag(inf2$cooksd, cutoff="internal")
inf3 <- hlm_influence(m1,level="zoo")
dotplot_diag(inf3$cooksd, cutoff="internal")
```


`r solend()`



`r qbegin(qcounter())`
Write a short description of the sample data.  


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


`r solend()`


`r qbegin(qcounter())`
Write a short explanation of your methods


::: {.callout-tip collapse="true"}
#### Hints

- explain transformations to the data prior to modelling
- explain the model(s) fitted and how they address the research aims
- describe any actions taken as a result of non-convergence
- what estimation method was used (ML, REML)? What software (including version and packages)? What optimiser? 
- if degrees of freedom were used, what kind? If not, how are inferences conducted?  
- what assumptions of the model were checked, and what was the outcome?


:::


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`


`r qbegin(qcounter())`
Write a short summary of your results, along with suitable visualisations and tables


::: {.callout-tip collapse="true"}
#### Hints

- if model comparisons were conducted, provide results and what they mean  
- present and interpret fixed effects estimates that are relevant to the research aim.  
- discuss random effect estimates - does the relevant effect vary a lot between clusters?  
- was the sensitivty of the results to influential observations/clusters assessed? if so, what did you find?


:::


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

`r solend()`



