---
title: "<b>Week 5: Individual Differences, part 1</b>"
subtitle: "Multivariate Statistics and Methodology using R (MSMR)<br><br> "
author: "Dan Mirman"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE)
options(digits = 4)
library(knitr)
library(kableExtra)
library(patchwork)

library(tidyverse)
library(broom)
library(broom.mixed)
library(gt)
library(lme4)
library(lmerTest)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  extra_css = list(".scroll-output" = list("height"="90%","overflow-y"="scroll"))
) 
```

# Individual differences

**Individual differences provide an additional level of analysis for understanding phenomena**

--

* At a group level, a treatment works better than a placebo, but why does it work better for some people than for others?
* People solve easy problems faster than hard problems, but why are some people a lot faster on the easy problems and other people only a little faster?

--

t-test and ANOVA methods treat individual differences as noise

Multilevel models provide two ways to quantify and analyse individual differences

---
# "External" individual differences can be added as fixed effects

Example: USA National Youth Survey longitudinal data on tolerance for deviant behaviour, exposure to deviant behaviour, and gender. Subset of 16 subjects printed in Willett (1997, Table 11.1). (`DeviantBehavior`)

`Tolerance` was measured by asking whether it is wrong for someone their age to: 

* cheat on tests
* purposely destroy property of others
* hit or threaten someone without reason
* use alcohol
* use marijuana
* sell hard drugs
* steal something worth less than $5
* steal something worth more than $50
* break into a building or vehicle to steal

---
# Tolerance for deviant behaviour

```{r}
load("./data/DeviantBehavior.RData")
summary(DeviantBehavior)
```

```{r echo=FALSE, fig.height=3.25, fig.width=5}
# The overall pattern (ignoring gender for now)
ggplot(DeviantBehavior, aes(Age, Tolerance, color=Gender)) + 
  stat_summary(fun=mean, geom="point", size=3) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) + 
  stat_summary(fun=mean, geom="line") +
  labs(x="Age", y="Tolerance for deviant behavior") + 
  theme_bw()
```

---
## Research question 1

**Does tolerance increase with age and is it modulated by gender?**

```{r, warning=FALSE}
m.base <- lmer(Tolerance ~ Age*Gender + (Age | SubjID),
                 contrasts=list(Gender="contr.sum"), data=DeviantBehavior)
```
```{r echo=FALSE}
gt(tidy(m.base))
```

Significant increase with age, no significant effects of gender

---
## Research question 2

**Is this modulated by exposure to deviant behavior?**

`Exposure` scale is not clear and other effects would be estimated at `Exposure=0`, which is not an attested value. 

If we center `Exposure`, its effects and the effects of other predictors will be easier to interpret.

```{r}
# center Exposure, but do not re-scale it
DeviantBehavior$Exposure.center <- scale(DeviantBehavior$Exposure, scale=FALSE)
```

--

Fit the full model

```{r}
m.exp1 <- lmer(Tolerance ~ Age*Gender*Exposure.center + (Age | SubjID), 
               data=DeviantBehavior, contrasts=list(Gender="contr.sum"))
```

---

```{r}
gt(tidy(m.exp1))
```

* Significant main effects: **Age** and **Exposure**
* Significant interactions: **Age-by-Exposure**, **Gender-by-Exposure**, **Age-by-Gender-by-Exposure**

---
# How to plot the three-way interaction?

The three-way Age-by-Gender-by-Exposure interaction is a relationship among four variables (Tolerance for deviant behaviour, Exposure to deviant behaviour, Age, and Gender), three of which are continuous variables. **This is hard to visualize.**

To make it easier to visualize, can split the `Exposure` into levels

--

Option 1: Median split
```{r}
DeviantBehavior$ExposureMedSpl <- ifelse(
  DeviantBehavior$Exposure >= median(DeviantBehavior$Exposure), "High", "Low")
```

---

```{r fig.height=3.5, fig.width=7}
ggplot(DeviantBehavior, aes(Age, Tolerance, color=ExposureMedSpl)) + 
  facet_wrap(~ Gender) + 
  stat_summary(fun=mean, geom="point", size=3) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) + 
  stat_summary(aes(y=fitted(m.exp1)), fun=mean, geom="line") +
  labs(x="Age", y="Tolerance for deviant behavior", 
       color="Exposure\n(Median Split)") + 
  theme_bw() 
```

Adolescents with higher Exposure to deviant behaviour tend to have increased Tolerance for deviant behaviour as they get older, and more so for males than females.

---
# How to plot the three-way interaction?

Option 2: tertile split provides a little more info than median split

```{r warning=FALSE}
# define break points
b <- quantile(DeviantBehavior$Exposure, probs=seq(0, 1, by=1/3)) 
# split continuous predictor and provide level labels
DeviantBehavior$Exposure3 <- cut(DeviantBehavior$Exposure, 
                                 breaks=b, include.lowest=T, 
                                 labels=c("Low", "Medium", "High"))
```

---
```{r fig.height=4, fig.width=8}
ggplot(DeviantBehavior, aes(Age, Tolerance, color=Exposure3)) + 
  facet_wrap(~ Gender) + 
  stat_summary(fun=mean, geom="point", size=3) + 
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) + 
  stat_summary(aes(y=fitted(m.exp1)), fun=mean, geom="line") +
  labs(x="Age", y="Tolerance for deviant behavior", 
       color="Exposure\n(Tertiles)") + 
  theme_bw() 
```

**Warning**: reviewers and readers may get confused about whether your *model* used continuous or categorical predictors, so you will need to be extra clear about this in your write-up.

---
# Key points

* Individual differences provide deeper insights into group-level phenomena
* Can assess them by adding them to the multilevel model
* Dividing a continuous variable into levels can help to visualise complex interactions