---
title: "<b>Week 2: Longitudinal Data Analysis using Multilevel Modeling</b>"
subtitle: "Multivariate Statistics and Methodology using R (MSMR)<br><br> "
author: "Dan Mirman"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE)
options(digits = 4)
library(knitr)
library(kableExtra)
library(patchwork)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  extra_css = list(".scroll-output" = list("height"="90%","overflow-y"="scroll"))
) 
```

# Longitudinal data are a natural application domain for MLM

* Longitudinal measurements are *nested* within subjects (by definition)
* Longitudinal measurements are related by a continuous variable, spacing can be uneven across participants, and data can be missing
    + These are problems rmANOVA
* Trajectories of longitudinal change can be nonlinear (we'll get to that next week)

--

## We've already seen some of this...

* Weight maintenance (Week 1 lab)
* Visual search (not longitudinal, but principles are the same)

---
# Another example

County-level percentage of adults who are physically active at recommended levels. Data from PHE for 2012-2015 (`PHE_MentalHealth.Rdata`)

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(effects)
load("./data/PHE_MentalHealth.Rdata")
str(mh_phe)
```

---
# Indicators
```{r}
gt::gt(unique(mh_phe[, 1:2]))
```

The indicator ID for percentage of physically active adults is 90275

---

```{r fig.width=8, fig.height=4, warning=FALSE}
filter(mh_phe, IndicatorID == 90275) %>%
  ggplot(aes(Year, Value, color = Region)) +
    stat_summary(fun = mean, geom="point") +
    geom_smooth(method = "lm", se=FALSE)
```

Looks like there are some interesting differences. We can ask two kinds of questions:

1.  Did the regions differ in their *baseline (2012)* percentage physically active adults?
2.  Did the regions differ in their *slopes of change* of percentage physically active adults?

---
# What is the baseline?

**Did the regions differ in their *baseline (2012)* percentage physically active adults?**

We can answer this question using the **intercept** coefficients

--

But those will be estimated at `Year = 0`, and the question is not about physical activity the year Jesus was born.

So we need to adjust the time variable so that 2012 corresponds to time 0 (and we can select just the physical activity values while we're at it):

```{r}
active_dat <- filter(mh_phe, IndicatorID == 90275) %>%
  mutate(Time = Year - 2012)
```

Now `Time` is a variable just like `Year`, but going from 0 to 3 instead of 2012 to 2015

---
# Fit the models

```{r}
# base model: just change over time
m <- lmer(Value ~ Time + (Time | County), data = active_dat, REML = F)
# add baseline differences between regions
m.0 <- lmer(Value ~ Time + Region + (Time | County), data = active_dat, REML = F)
# add slope differences between regions
m.1 <- lmer(Value ~ Time * Region + (Time | County), data = active_dat, REML = F)
```

---
# Compare models

```{r}
anova(m, m.0, m.1)
```

Regions differed in baseline percentage of physically active adults ( $\chi^2(8) = 39.3, p < 0.0001$ ), but there not in terms of the slope of change ( $\chi^2(8) = 9.86, p = 0.28$ ).

---
# Visualize effects, Option 1: model-estimated percentages

```{r fig.width=8, fig.height=4}
ef <- as.data.frame(effect("Time:Region", m.1))
ggplot(ef, aes(Time, fit, color=Region)) + 
  geom_line() +
  theme_bw() + scale_color_brewer(palette = "Set1")
```

---
# Visualize effects, Option 2: estimated parameters

```{r echo=FALSE}
est <- broom::tidy(m.1) %>% 
  filter(group == "fixed") %>% 
  mutate(Type = ifelse(str_detect(term, "Time"), "Slope", "Baseline")) %>%
  arrange(Type, term) %>% 
  mutate(Region = rep(sort(unique(active_dat$Region)), 2))
# need to adjust non-reference estimates 
est$estimate[2:9] <- est$estimate[2:9] + est$estimate[1] # adjust baselines using reference intercept
est$estimate[11:18] <- est$estimate[11:18] + est$estimate[10] # adjust slopes using reference slope
# order the regions by baseline physical activity
est <- est %>% mutate(Region = fct_reorder(Region, estimate, .fun=sum))
```

[Side note: it took me a while to work out the data wrangling required to get the coefficients into a convenient form for plotting]

```{r fig.width=9, fig.height=4}
ggplot(est, aes(Region, estimate, ymin=(estimate-std.error), ymax=(estimate+std.error))) +
  facet_wrap(~ Type, scales="free_x") +
  geom_pointrange() + coord_flip()
```

