---
title: "<b>Multilevel Model Checklist (sort of)</b>"
subtitle: "Multivariate Statistics and Methodology using R<br><br> "
author: "MSMR TEAM"
institute: "Department of Psychology<br/>The University of Edinburgh"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: jk_libs/libs
    css: 
      - xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(digits=4,scipen=2)
options(knitr.table.format="html")
library(xaringanthemer)
style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
  # base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  extra_css = list(".scroll-output" = list("height"="90%","overflow-y"="scroll"))
) 
knitr::opts_chunk$set(
  dev = "png",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
themedapr3 = function(){
  theme_minimal() + 
    theme(text = element_text(size=20))
}
```

```{r echo=FALSE}
library(tidyverse)
library(lme4)
library(lmeresampler)
library(HLMdiag)
library(broom.mixed)
```


---
class: inverse, center, middle

<h2>Part 1: Model Specification</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Analysis</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Reporting</h2>

---
# Outcome and Fixed Effects  

.br2.f4.gray.bg-white[
(g)lmer(**outcome** ~ fixed effects + (random effects | grouping structure), family = error distribution)
]

.pull-left[
- __What are we interested in explaining/predicting?__  

]

---
count:false
# Outcome and Fixed Effects  

.br2.f4.gray.bg-white[
**(g)lmer**(outcome ~ fixed effects + (random effects | grouping structure), **family = error distribution**)
]

.pull-left[

- What are we interested in explaining/predicting?  

  - __How is this measured?__

]

---
count:false
# Outcome and Fixed Effects  

.br2.f4.gray.bg-white[
(g)lmer(outcome ~ **fixed effects** + (random effects | grouping structure), family = error distribution)
]

.pull-left[

- What are we interested in explaining/predicting?  

  - How is this measured?

- __What variables are we interested in explaining this by?__  

]

---
# The Grouping Structure  

.br2.f4.gray.bg-white[
(g)lmer(outcome ~ fixed effects + (random effects | grouping structure), family = error distribution)
]

.pull-left[
- In what different ways can we group our data?   

]

---
count:false
# The Grouping Structure  

.br2.f4.gray.bg-white[
(g)lmer(outcome ~ **fixed effects** + (random effects | grouping structure), family = error distribution)
]

.pull-left[
- In what different ways can we group our data?   

- **Of the different ways we can group our data, which groupings are of specific inferential interest?**  

- Of the different ways we can group our data, which groupings do we think of as a random sample from a general population of groups? 

]

---
count:false
# The Grouping Structure  

.br2.f4.gray.bg-white[
(g)lmer(outcome ~ fixed effects + (random effects | **grouping structure**), family = error distribution)
]

.pull-left[
- In what different ways can we group our data?   

- Of the different ways we can group our data, which groupings are of specific inferential interest?  

- **Of the different ways we can group our data, which groupings do we think of as a random sample from a general population of groups?** 

]


---
# The Grouping Structure  

.br2.f4.gray.bg-white[
(g)lmer(outcome ~ fixed effects + (random effects | **grouping1 / grouping2**), family = error distribution)
]

.pull-left[
- In what different ways can we group our data?   

- Of the different ways we can group our data, which groupings are of specific inferential interest?  

- Of the different ways we can group our data, which groupings do we think of as a random sample from a general population of groups? 

  - Is there more than one grouping of this sort?
      - __are these groupings nested? (and are the labels unique?)__  
]

---
# The Grouping Structure  

.br2.f4.gray.bg-white[
(g)lmer(outcome ~ fixed effects + (random effects | **grouping1**) + (random effects | **grouping2**), family = error distribution)
]

.pull-left[
- In what different ways can we group our data?   

- Of the different ways we can group our data, which groupings are of specific inferential interest?  

- Of the different ways we can group our data, which groupings do we think of as a random sample from a general population of groups? 

  - Is there more than one grouping of this sort?  
      - are these groupings nested? (and are the labels unique?)  
      - __are the groupings crossed?__  
  
]

---
# The Grouping Structure  

.br2.f4.gray.bg-white[
(g)lmer(outcome ~ fixed effects + (random effects | **grouping structure**), family = error distribution)
]

.pull-left[
- In what different ways can we group our data?   

- Of the different ways we can group our data, which groupings are of specific inferential interest?  

- Of the different ways we can group our data, which groupings do we think of as a random sample from a general population of groups? 

  - Is there more than one grouping of this sort?  
      - are these groupings nested? (and are the labels unique?)   
      - are the groupings crossed?  
  - For each level, how many groups have we sampled? (If small number, consider including it as a fixed effect instead). 
]


---
# Random Intercepts and Slopes

.br2.f4.gray.bg-white[
(g)lmer(outcome ~ fixed effects + (**random effects** | grouping structure), family = error distribution)
]

.pull-left[

- Which of our fixed effects can vary for our random groups?  

  - "does a single group have multiple values for $x$?"  
  
  - "for the data from only one of our groups, can we estimate $y \sim x$?"

]


---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Model Specification</h2>
<h2>Part 2: Analysis</h2>
<h2 style="text-align: left;opacity:0.3;">Part 3: Reporting</h2>


---
# Model issues

- Keep an eye out for convergence issues & singular fits  



  - Adjust optimisers, try all available optimisers (the `allFit()` function).  
  - Consider simplifying your random effect structure.
  - Look out for:  
      - random effects with very small variance estimates (scale of outcome variable is relevant here)
      - very strong correlations between random effects 



- Always plot your residuals and your random effects.


---
# Inference

.pull-left[
## Tests

- Model comparison
  - Remember to be aware when you are fitting models with `REML=FALSE`

- Parameter estimates


]
.pull-right[
## Methods

- df approximations
  - load `library(lmerTest)`
  - `pbkrtest::KRmodcomp()`
  
- Likelihood Ratio Tests  
  `anova(model1, model2, ...)`
- Bootstrap
  - parametric bootstrap
    `pbkrtest::PBmodcomp()` and `confint(method="boot")`
  - case bootstrap
    `lmeresampler::bootstrap(model, type = "case", resample = c(....))`
]


---
class: inverse, center, middle

<h2 style="text-align: left;opacity:0.3;">Part 1: Model Specification</h2>
<h2 style="text-align: left;opacity:0.3;">Part 2: Analysis</h2>
<h2>Part 3: Reporting</h2>

---
# Reporting the analysis process

- Data cleaning outlier/data removal, transformations _prior to_ analysis.  

- Unplanned transformations and data removal which are carried out in order to meet assumptions.  

--

- Specify all fixed effects (explanatory variables & covariates).  
Link them to explicitly stated research questions/hypotheses. 

- Explicitly state the hierarchical structure of the data and of the model.  
Specify random effects according to the sampling units (schools/children etc) with which they interact. 

--

- State the software packages and versions used to fit the model(s), along with the estimation method (ML/REML) and optimiser used.   

- If a proposed model fails to converge, clearly specify the procedure used to obtain converging model.  

--

- State clearly the relevant test/comparison/parameter estimate of interest.  
Link to explicitly stated research questions/hypotheses.  
  - Any model comparisons should be clearly stated so that the reader understands the structure of _both_ models being compared.  
- Specify the method you plan to use to conduct inference (e.g. LRT, bootstrap)

---
# Reporting results from the model(s)

Information to include:

- all parameter estimates for fixed effects.  

    - coefficients
    - standard errors and/or confidence intervals
    - associated test statistics and p-values (if used)  
    
- random effects  

  - standard deviation and/or variance for each random effect
  - correlations/covariances if modelled   
  - residual variance/standard deviation
  
- some measure of model fit (marginal/conditional $R^2$)?  

<br>
`sjPlot::tab_model()` gets you a lot of the way there! 

---
# Visualising the model(s)  

- Think about your questions  

.pull-left[
```{r echo=FALSE, fig.asp=.8}
ggplot()+
  geom_text(label="?",aes(x=1,y=1), size=40)+
  scale_x_continuous("Some explanatory variable of interest",labels=NULL)+
  scale_y_continuous("Outcome",labels=NULL)+
  theme_classic()
```
]

--

.pull-right[

- `sjPlot::plot_model()` for quick easy plotting (less control)  

- `as.data.frame(effects::effect(term = ??, mod = model))` for a dataframe of linear predictions you can use to plot.  

- `broom.mixed::augment(model)` for the model fitted values (good for plotting group-level lines etc). 

- `lme4::dotplot.ranef.mer(model)` for caterpillar plots of random effects

]

---
class: inverse, center, middle, animated, rotateInDownLeft

# End

