---
title: "Logistic Multilevel Models"
bibliography: references.bib
biblio-style: apalike
link-citations: yes
params: 
    SHOW_SOLS: TRUE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
source('assets/setup.R')
library(patchwork)
library(tidyverse)
library(lme4)
library(broom.mixed)
library(effects)
```

:::blue
**Preliminaries**  
 
1. Open Rstudio, create a new R Script or RMarkdown document (whichever you prefer working with) and give it a title for this week. 

2. Load any of the packages you think you might be using today.  
By now you may have started get into a sort of routine with R, and you might know what functions you like, and which you don't. Because there are so many alternatives for doing the same thing, the packages and functions you use are very much up to you.   
If I'm planning on doing some multi-level modelling, I will tend to load these by default at the top: 
    ```{r}
    library(tidyverse) # for everything
    library(lme4) # for fitting models
    library(broom.mixed) # for tidying model output
    ```
    
:::


# Exercises

:::rtip
__lmer() >> glmer()__

Remember how we simply used `glm()` and could specify the `family = "binomial"` in order to fit a logistic regression? Well it's much the same thing for multi-level models! 

+ Gaussian model: 
  - `lmer(y ~ x1 + x2 + (1 | g), data = data)`  
+ Binomial model:  
  - `glmer(y ~ x1 + x2 + (1 | g), data = data, family = binomial(link='logit'))`<br>or
  - `glmer(y ~ x1 + x2 + (1 | g), data = data, family = "binomial")`<br>or
  - `glmer(y ~ x1 + x2 + (1 | g), data = data, family = binomial)`
    
For more on Binary vs Binomial, see this [in depth section](zz_binary_binomial.html).

:::

:::frame
__Novel Word Learning: Data Codebook__

```{r}
load(url("https://uoepsy.github.io/msmr/data/nwl.RData"))
```

In the `nwl` data set (accessed using the code above), participants with aphasia are separated into two groups based on the general location of their brain lesion: anterior vs. posterior. There is data on the numbers of correct and incorrect responses participants gave in each of a series of experimental blocks. There were 7 learning blocks, immediately followed by a test. Finally, participants also completed a follow-up test.  
<br>
Data were also collect from healthy controls. 
<br>
Figure \@ref(fig:nwl-fig) shows the differences between groups in the average proportion of correct responses at each point in time (i.e., each block, test, and follow-up)

```{r nwl-fig, echo=FALSE, fig.cap="Differences between groups in the average proportion of correct responses at each block"}
load(url("https://uoepsy.github.io/msmr/data/nwl.RData"))
ggplot(filter(nwl, !is.na(lesion_location)), aes(block, PropCorrect, 
                                            color=lesion_location, 
                                            shape=lesion_location)) +
  #geom_line(aes(group=ID),alpha=.2) + 
  stat_summary(fun.data=mean_se, geom="pointrange") + 
  stat_summary(data=filter(nwl, !is.na(lesion_location), block <= 7), 
                           fun=mean, geom="line") + 
  geom_hline(yintercept=0.5, linetype="dashed") + 
  geom_vline(xintercept=c(7.5, 8.5), linetype="dashed") + 
  scale_x_continuous(breaks=1:9, labels=c(1:7, "Test", "Follow-Up")) + 
  theme_bw(base_size=10) + 
  labs(x="Block", y="Proportion Correct", shape="Lesion\nLocation", color="Lesion\nLocation")
```


```{r echo=FALSE}
data.frame(
  variable = names(nwl),
  description = c("Whether participant is a stroke patient ('patient') or a healthy control ('control')", "Location of brain lesion: anterior vs posterior","Experimental block (1-9). Blocks 1-7 were learning blocks, immediately followed by a test in block 8. Block 9 was a follow-up test at a later point","Proportion of 30 responses in a given block that the participant got correct","Number of responses (out of 30) in a given block that the participant got correct","Number of responses (out of 30) in a given block that the participant got incorrect","Participant Identifier","Experimental phase, corresponding to experimental block(s): 'Learning', 'Immediate','Follow-up'")
) %>% knitr::kable(.)
```



:::


`r qbegin("A1")`
Load the data. Take a look around. Any missing values? Can you think of why?  
`r qend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
load(url("https://uoepsy.github.io/msmr/data/nwl.RData"))
summary(nwl)
```
The only missing vales are in the lesion location, and it's probably because the healthy controls don't have any lesions. 

`r solend()`


`r qbegin("A2")`
Our broader research aim today is to compare two groups (those with anterior vs. posterior lesions) with respect to their accuracy of responses over the course of the study.  

1. What is our outcome? 

*Hint:* Think carefully: there might be several variables which either fully or partly express the information we are considering the "outcome" here. 
`r qend()`
`r solbegin(show=TRUE, toggle=params$TOGGLE)`
1. The outcome here is (in words) "the proportion of correct answers in each block". This makes it tempting to look straight to the variable called `PropCorrect`. Unfortunately, this is encoded as a proportion (i.e., is bounded by 0 and 1), and of the models we have learned about so far, we don't definitely have the necessary tools to model this.  
    - linear regression = continuous *unbounded* outcome variable  
    - logistic regression = binomial, expressed as `0` or `1` if the number of trials per observation is 1, and expressed as `cbind(num_successes, num_failures)` if the number of trials per observation is >1. 
`r solend()`

`r qbegin("A3")`
> **Research Question 1:**  
> Is the learning rate (training blocks) different between the two lesion location groups?

**Hints**: 

- Do we want `cbind(num_successes, num_failures)`?

- Make sure we're running models on only the data we are actually interested in.  

- We could use model comparison via likelihood ratio tests (using `anova(model1, model2, model3, ...)`. For this question, we could compare:

    - A model with just the change over the sequence of blocks
    - A model with the change over the sequence of blocks *and* an overall difference between groups
    - A model with groups differing with respect to their change over the sequence of blocks  

- What about the random effects part?  
    
    1. What are our observations grouped by? 
    2. What variables can vary within these groups? 
    3. What do you want your model to allow to vary within these groups?

`r qend()`

`r optbegin("Suggested answers to the hints if you don't know where to start", olabel=FALSE, toggle=params$TOGGLE)`

- Make sure we're running models on only the data we are actually interested in. 
    - We want only the learning blocks, and none of the healthy controls.
    - You might want to store this data in a separate object, but in the code for the solutio we will just use `filter()` *inside* the `glmer()`.   
  
- A model with just the change over the sequence of blocks:
    - **outcome ~ block**
- A model with the change over the sequence of blocks *and* an overall difference between groups:
    - **outcome ~ block + lesion_location**
- A model with groups differing with respect to their change *over the sequence of blocks:
    - **outcome ~ block * lesion_location**
    
- What are our observations grouped by? 
    - repeated measures by-participant. i.e., the `ID` variable
- What variables can vary within these groups? 
    - `Block` and `Phase`. Be careful though - you can create the `Phase` variable out of the `Block` variable, so really this is just one piece of information, encoded differently in two variables. 
    - The other variables (`lesion_location` and `group`) do **not** vary for each ID. Lesions don't suddenly change where they are located, nor do participants swap between being a patient vs a control (we don't need the group variable anyway as we are excluding the controls).  
What do you want your model to allow to vary within these groups?
    - Do you think the change over the course of the blocks is **the same** for everybody? Or do you think it varies? Is this variation important to think about in terms of your research question?   
    
`r optend()`

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
m.base <- glmer(cbind(NumCorrect, NumError) ~ block + (block | ID), 
                data = filter(nwl, block < 8, !is.na(lesion_location)),
                family=binomial)
m.loc0 <- glmer(cbind(NumCorrect, NumError) ~ block + lesion_location + (block | ID), 
                data=filter(nwl, block < 8, !is.na(lesion_location)),
                family=binomial)
m.loc1 <- glmer(cbind(NumCorrect, NumError) ~ block * lesion_location + (block | ID), 
                data=filter(nwl, block < 8, !is.na(lesion_location)),
                family=binomial)
#summary(m.loc1)
anova(m.base, m.loc0, m.loc1, test="Chisq")
```
:::int
No significant difference in learning rate between groups ($\chi^2(1)=2.2, p = 0.138$).
:::

`r solend()`

`r qbegin("A4")`
> **Research Question 2**  
> In the testing phase, does performance on the immediate test differ between lesion location groups, and does their retention from immediate to follow-up test differ?

Let's try a different approach to this. Instead of fitting various models and comparing them via likelihood ratio tests, just fit the one model which could answer both parts of the question above.  

**Hints:**
    
- This might required a bit more data-wrangling before hand. Think about the order of your factor levels
(alphabetically speaking, "Follow-up" comes before "Immediate")!

`r qend()`
`r solbegin(show=params$SHOW, toggle=params$TOGGLE)`
```{r}
nwl_test <- filter(nwl, block > 7, !is.na(lesion_location)) %>%
  mutate(
    Phase = fct_relevel(factor(Phase),"Immediate")
  )

m.recall.loc <- glmer(cbind(NumCorrect, NumError) ~ Phase * lesion_location + (Phase | ID), 
                  nwl_test, family="binomial")

summary(m.recall.loc)
```
`r solend()`

:::frame
__Interpreting coefficients in logistic regression__   
Take some time to remind yourself from USMR of the interpretation of logistic regression coefficients.  

The interpretation of the fixed effects of a logistic multilevel model is not very different.  
We can obtain the fixed effects from our model by using:  

- `fixef(model)`
- `summary(model)$coefficients`  
- `coef(summary(model))`
- `tidy(model)` from the **broom.mixed** package  
- (there are probably more ways, but I can't think of them right now!)

It's just that for multilevel models, we can model by-cluster random variation around these effects.  
:::

`r qbegin("A5")`

1. In `family = binomial(link='logit')`. What function is used to relate the linear predictors in the model to the expected value of the response variable?  
2. How do we convert this into something more interpretable?  

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
 
1. The link function is the `logit`, or log-odds (other link functions are available).

2. To convert log-odds to odds, we can use `exp()`, to get odds and odds ratios.  

`r solend()`

`r qbegin("A6")`
Make sure you pay attention to trying to interpret each fixed effect from your models.  
These can be difficult, especially when it's logistic, and especially when there are interactions.  

- What is the increase in the odds of answering correctly in the immediate test for someone with a posterior legion compared to someone with an anterior legion?  

<!-- `r optbegin("Optional help: Our Solution to A4", olabel=F, toggle=params$TOGGLE)` -->
<!-- ```{r eval=F} -->
<!-- nwl_test <- filter(nwl, block > 7, !is.na(lesion_location)) %>% -->
<!--   mutate( -->
<!--     Phase = fct_relevel(factor(Phase),"Immediate") -->
<!--   ) -->

<!-- m.recall.loc <- glmer(cbind(NumCorrect, NumError) ~ Phase * lesion_location + (Phase | ID),  -->
<!--                   nwl_test, family="binomial") -->
<!-- ``` -->
<!-- `r optend()` -->

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
* `(Intercept)` ==> Anterior lesion group performance in immediate test. This is the log-odds of them answering correctly in the immediate test. 
* `PhaseFollow-up`  ==> Change in performance (for the anterior lesion group) from immediate to follow-up test. 
* `lesion_locationposterior` ==> Posterior lesion group performance in immediate test relative to anterior lesion group performance in immediate test
* `PhaseFollow-up:lesion_locationposterior` ==> Change in performance from immediate to follow-up test, posterior lesion group relative to anterior lesion group

```{r echo=FALSE}
exp(fixef(m.recall.loc))[3]
```

:::int
Those with posterior lesions have `r round(exp(fixef(m.recall.loc))[3],2)` times the odds of answering correctly in the immediate test compared to someone with an anterior lesion. 
:::

`r solend()`

`r qbegin("OPTIONAL: A7", qlabel=F)`
Recreate the visualisation in Figure \@ref(fig:nwl-fig2).

```{r nwl-fig2, echo=FALSE, fig.cap="Differences between groups in the average proportion of correct responses at each block"}
load(url("https://uoepsy.github.io/msmr/data/nwl.RData"))
ggplot(filter(nwl, !is.na(lesion_location)), aes(block, PropCorrect, 
                                                 color=lesion_location, 
                                                 shape=lesion_location)) +
    #geom_line(aes(group=ID),alpha=.2) + 
    stat_summary(fun.data=mean_se, geom="pointrange") + 
    stat_summary(data=filter(nwl, !is.na(lesion_location), block <= 7), 
                 fun=mean, geom="line") + 
    geom_hline(yintercept=0.5, linetype="dashed") + 
    geom_vline(xintercept=c(7.5, 8.5), linetype="dashed") + 
    scale_x_continuous(breaks=1:9, 
                       labels=c(1:7, "Test", "Follow-Up")) + 
    theme_bw(base_size=10) + 
    labs(x="Block", y="Proportion Correct", 
         shape="Lesion\nLocation", color="Lesion\nLocation")
```

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
ggplot(filter(nwl, !is.na(lesion_location)), aes(block, PropCorrect, 
                                                 color=lesion_location, 
                                                 shape=lesion_location)) +
    #geom_line(aes(group=ID),alpha=.2) + 
    stat_summary(fun.data=mean_se, geom="pointrange") + 
    stat_summary(data=filter(nwl, !is.na(lesion_location), block <= 7), 
                 fun=mean, geom="line") + 
    geom_hline(yintercept=0.5, linetype="dashed") + 
    geom_vline(xintercept=c(7.5, 8.5), linetype="dashed") + 
    scale_x_continuous(breaks=1:9, 
                       labels=c(1:7, "Test", "Follow-Up")) + 
    theme_bw(base_size=10) + 
    labs(x="Block", y="Proportion Correct", 
         shape="Lesion\nLocation", color="Lesion\nLocation")
```
`r solend()`

:::frame
__Coding Schemes for Categorical Predictors__  
  
Remember that categorical predictors get inputted into our model as a series of variables which are 0s and 1s? Previously the variable in our model for lesion location was actually being coded with 0 representing one level, and 1 representing the other. This is known as "treatment coding".  

There are lots of other ways we might encode our categorical predictors. One common approach is to use "effects coding". In the case where we have a binary predictor, this makes zero the mid-point between the two  - i.e., the overall mean (this is a bit like mean-centering a continuous predictor).  

If we recall that the intercept is "when all predictors are zero", and that when we have an interaction `Y~A+B+A*B` in our model, the individual coefficients for `A` and `B` are estimated "when the other variable is zero", then we can start to understand how these different ways of encoding our categorical predictors can change what we are getting out of our model. (Note, they don't actually change anything about the model fit, but they change what information we are estimating from the model).  

```{r echo=FALSE}
filter(nwl, !is.na(lesion_location)) %>%
    group_by(lesion_location) %>%
    summarise(m = mean(PropCorrect)) %>% pull(m) -> lmeans
filter(nwl, !is.na(lesion_location)) %>%
    mutate(isPost = 1*(lesion_location=="posterior")) %>%
    ggplot(.,aes(x=isPost,y=PropCorrect))+
    geom_jitter(height=0,width=.04,alpha=.2)+
    stat_summary(geom="pointrange")+
    geom_segment(x=0, xend=1, y=lmeans[1],yend=lmeans[2],col="blue")+
    labs(x="lesion_location",title="treatment coding", subtitle="contr.treatment")+
    scale_x_continuous(breaks=c(0,1),labels=c("0\n(anterior)","1\n(posterior)"))+
    geom_text(x=0.15,y=.45,label="intercept",col="red")+
    geom_curve(x=0.09,xend=0,y=0.45,yend=lmeans[1],curvature = -0.1,arrow = arrow(length = unit(0.3, "cm")),col="red")+
    geom_segment(x=0,xend=1,y=lmeans[1],yend=lmeans[1],lty="dotted",col="red")+
    geom_curve(x=1,xend=1,y=lmeans[1],yend=lmeans[2],curvature = 0,arrow = arrow(length = unit(0.3, "cm")),col="red") -> tcod

filter(nwl, !is.na(lesion_location)) %>%
    mutate(isPost = 1*(lesion_location=="posterior"),
           isPost = ifelse(isPost==0,-1,isPost)) %>%
    ggplot(.,aes(x=isPost,y=PropCorrect))+
    geom_jitter(height=0,width=.04,alpha=.2)+
    stat_summary(geom="pointrange")+
    geom_segment(x=0, xend=1, y=lmeans[2]-(0.5*diff(lmeans)),yend=lmeans[2],col="blue",lty="dashed")+
    labs(x="lesion_location",title="effects coding",subtitle="contr.sum")+
    scale_x_continuous(breaks=c(-1,0,1),labels=c("-1\n(anterior)","0","1\n(posterior)"))+ 
    geom_text(x=0.15,y=.66,label="intercept",col="red")+
    geom_curve(x=0.09,xend=0,y=0.65,yend=lmeans[1]+(0.5*diff(lmeans)),curvature = -0.1,arrow = arrow(length = unit(0.3, "cm")),col="red")+
    geom_segment(x=0,xend=1,y=lmeans[1]+(0.5*diff(lmeans)),yend=lmeans[1]+(0.5*diff(lmeans)),lty="dotted",col="red")+
    geom_curve(x=1,xend=1,y=lmeans[1]+(0.5*diff(lmeans)),yend=lmeans[2],curvature = 0,arrow = arrow(length = unit(0.3, "cm")),col="red") -> efcod
tcod + efcod
```

If you want to understand this a little more, take a look at this [in depth section](zz_quickcontrasts.html). 

:::

`r qbegin("A8")`
This code is that we used to answer question A4 above, only we have edited it to change lesion location to be fitted with "effects coding".  

```{r echo=TRUE}
nwl_test <- filter(nwl, block > 7, !is.na(lesion_location)) %>%
    mutate(
        Phase = fct_relevel(factor(Phase),"Immediate")
    )

m.recall.loc.effcoding <- glmer(cbind(NumCorrect, NumError) ~ Phase * lesion_location + (Phase | ID), 
                                contrasts = list(lesion_location = "contr.sum"),
                                data = nwl_test, family="binomial")
```


The interpretation of this is going to get pretty tricky - we have a logistic regression, and we have different coding scheme for our categorical predictor, and we have an interaction.. &#x1f92f;  

Can you work out the interpretation of the fixed effects estimates?  

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

* `(Intercept)` ==> Overall performance in immediate test. This is the overall log-odds of answering correctly in the immediate test. 
* `PhaseFollow-up`  ==> Average change in performance from immediate to follow-up test. 
* `lesion_location1` ==> Anterior lesion group performance in immediate test relative to *overall average* performance in immediate test
* `PhaseFollow-up:lesion_location1` ==> Change in performance from immediate to follow-up test, anterior lesion group relative to overall average
  
  
**???**  
How do we know that `lesion_location1` is the *anterior* and not the *posterior* lesion group? 
We need to check the what the contrasts look like:  
```{r}
contrasts(nwl_test$lesion_location) <- "contr.sum"
contrasts(nwl_test$lesion_location)
```
Because there are only two levels to this variable, the estimate will simply flip sign (positive/negative) depending on which way the contrast is leveled.  

`r optbegin("Optional: I liked my coefficients being named properly", olabel=FALSE, toggle=params$TOGGLE)`

```{r}
colnames(contrasts(nwl_test$lesion_location)) <- "PeppaPig"

contrasts(nwl_test$lesion_location)

modeltest <- glmer(cbind(NumCorrect, NumError) ~ Phase * lesion_location + (Phase | ID),
                   nwl_test, family="binomial")
summary(modeltest)$coefficients
```
`r optend()`

`r solend()`


:::frame
__Hang on.. p-values are back?!__ 

We noted at the end of last week that we don't have p-values for `lmer()`^[We also didn't get $R^2$. Check out [http://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#how-do-i-compute-a-coefficient-of-determination-r2-or-an-analogue-for-glmms](http://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#how-do-i-compute-a-coefficient-of-determination-r2-or-an-analogue-for-glmms)], but you might notice that we do now have them when we've fitted a model with `glmer()`?

The reason is partly just one of convention. There is a standard practice for determining the statistical significance of parameters in the generalised linear model, relying on asymptotic Wald tests which evaluate differences in log-likelihood.  
:::


# Reading: Inference in MLM  

In USMR, we fitted various simple linear models using `lm()`, and we could get out either a table of coefficients (for example, Table \@ref(tab:lmcoeftab)) or a table of the reduction in sums of squared residuals (for example, Table \@ref(tab:lmanovtab)). In both of these cases we had a nice set of p-values for us to look at (in R, we often find the p-values in the column named something like `Pr(>F)`).  

:::statbox
__A quick reminder:__ a p-value is the probability of observing results as or more extreme than the obtained ones, _if the data were really generated by the hypothesised model or process_).  
:::


```{r lmcoeftab, echo=FALSE}
mwdata = read_csv(file = "https://uoepsy.github.io/data/wellbeing.csv")
model2 <- lm(wellbeing ~ 1 + outdoor_time + social_int, data = mwdata)
summary(model2)$coefficients %>% knitr::kable(caption="t-tests for slope coefficients for predictors in a linear model")
```

```{r lmanovtab, echo=FALSE}
anova(model2) %>% knitr::kable(caption="F-tests for predictors in a linear model")
```


We could get p-values for these models because we know what the distribution of test statistics of interest (e.g. $F$-statistics and $t$-statistics), will look like if the null hypothesis were true (i.e., we can describe what they look like in terms of $F$ and $t$ distributions with specific degrees of freedom (look back to the [USMR materials for a formula](https://uoepsy.github.io/usmr/2223/labs/07a_slr.html#the-f-statistic)).  

Unfortunately, the same is not true of multilevel models. For the multilevel model, we can think of it as having residuals at multiple levels: we have the random variation of clusters around the fixed effects, and then random variation of observations around the cluster-level effects (more on this in Week 4).  
In the rare occasion that you have a perfectly balanced experimental design, then ratios of sums of squares for multi-level models follow an $F$-distribution, in which we know the numerator and denominator degrees of freedom (this means we can work out the degrees of freedom for the $t$-test of our fixed effect parameters). Unfortunately, in the real world where things are not often perfectly balanced, determining the denominator degrees of freedom becomes unclear.

Last week, we mentioned a couple of approaches that we might take for drawing inferences, finding ways to compute p-values or confidence intervals.^[It's always important to note that aiming for a p-value or confidence interval to make a dichotomous decision is only _one approach_ to doing statistics. If you continue to develop your learning of statistics after this course (which we hope you will!), then you will find that there are other schools of thought that bring different benefits.]  We're going to now extend this to include a few more, and discuss the strengths and limitations of the different approaches.  


:::frame
__For these examples... __  

For the following examples, we're going to return to our dataset of various toys, and we are going to be concerned with whether practice (the `hrs_week` variable) is associated with changes in reading ages (`R_AGE` variable).  
To accommodate for the clustered nature of the data, we are going to fit a model with both intercepts and slopes varying by toy-type.  
```{r}
toys_read <- read_csv("https://uoepsy.github.io/data/toyexample.csv")
full_model <- lmer(R_AGE ~ hrs_week + (1 + hrs_week | toy_type), data = toys_read)
summary(full_model)
```
  
:::

`r optbegin("Use a normal approximation (not advisable)", olabel=F,toggle=params$TOGGLE)`
Remember that the $t$ distribution starts to look more and more like the $z$ ("normal") distribution when degrees of freedom increase? We could just assume we have infinite degrees of freedom in our test statistics, and pretend that the $t$-values we get are actually $z$-values. This is "anti-conservative" in as much as it is not a very cautious approach, and we are likely to have a higher false positive rate (e.g. more chance of saying "there __is__ an effect!" when there actually isn't.)
```{r}
coefs <- as.data.frame(summary(full_model)$coefficients)
coefs$p.z <- 2 * (1 - pnorm(abs(coefs[,3])))
coefs
```
`r optend()`

`r optbegin("Satterthwaite df approximation", olabel=F,toggle=params$TOGGLE)`
There have been a couple of methods proposed to estimate the degrees of freedom in order to provide a better approximation to the null distribution of our tests. The way the Satterthwaite method has been implemented in R will just add a column for p-values to your `summary(model)` output).  
  
Load the __lmerTest__ package, refit the model, and voila!  
```{r}
library(lmerTest)
full_model <- lmer(R_AGE ~ hrs_week + (1 + hrs_week | toy_type), data = toys_read)
summary(full_model)$coefficients
```

:::int
__Reporting__  


To account for the extra uncertainty brought by the inclusion of random effects in the model, the degrees of freedom in the coefficients tests have been corrected via Satterthwaite's method.   
...   
...   
Weekly hours of reading practice was associated with increased reading age ($\beta = 1.14,\ SE = 0.30,\ t(17.52^*) = 3.87,\ p = .001$).

:::

__Note:__ if you have the __lmerTest__ package loaded, then all the models you fit with `lmer()` will show p-values! If you want to stop this, then you will have to detach/unload the package, and refit the model.  
```{r}
detach("package:lmerTest", unload=TRUE)
```

`r optend()`

`r optbegin("Kenward Rogers df approximations", olabel=F,toggle=params$TOGGLE)`
The Kenward-Rogers approach is slightly more conservative than the Satterthwaite method, and has been implemented for model comparison between a full model and a restricted model (a model without the parameter of interest), using the KR adjustment for the denominator degrees of freedom in the $F$-test.  
For this, models must be fitted with REML, __not__ ML. The function `KRmodcomp()` will take care of this and re-fit them for you.   
```{r}
library(pbkrtest)
restricted_model <- lmer(R_AGE ~ 1 + (1 + hrs_week | toy_type), data = toys_read)
full_model <- lmer(R_AGE ~ hrs_week + (1 + hrs_week | toy_type), data = toys_read)
KRmodcomp(full_model, restricted_model)
```

:::int
__Reporting__  

To account for the extra uncertainty brought by the inclusion of random effects in the model, the denominator degrees of freedom in have been corrected via Kenward-Rogers' method.  
...  
...  
Inclusion of weekly hours of reading practice as a predictor was associated with an improvement in model fit ($F(1,17.74^*) = 14.64,\ p = .001$).
:::
`r optend()`

`r optbegin("Likelihood Ratio Test (LRT)", olabel=F,toggle=params$TOGGLE)`
Conduct a model comparison between your model and a restricted model (a model without the parameter of interest), evaluating the change in log-likelihood.

:::statbox
__Likelihood__  

"likelihood" is a function that associates to a parameter the probability (or probability density) of observing the given sample data.  

The intuition behind likelihood:  

1. I toss a coin 10 time and observed 8 Heads.  
2. We can think of a 'model' of the process that governs the coin's behaviour in terms of just one number: a parameter that indicates the probability of the coin landing on heads.  
I have two models: 
  - Model 1: The coin will land on heads 20% of the time. $P(Heads)=0.2$  
  - Model 2: The coin will land on heads 70% of the time. $P(Heads)=0.7$  
3. Given the data I observe (see 1, above), we can (hopefully) intuit that Model 2 is more likely than Model 1. 


For a (slightly) more detailed explanation, see [here](zz_lvp.html).  

:::

This method assumes that the ratio of two likelihoods will (as sample size increases) become closer to being $\chi^2$ distributed, and so may be unreliable for small samples.  

Models must be fitted with ML, __not__ REML. The function `anova()` will re-fit them for you.   
```{r message=T,warning=T}
restricted_model <- lmer(R_AGE ~ 1 + (1 + hrs_week | toy_type), data = toys_read)
full_model <- lmer(R_AGE ~ hrs_week + (1 + hrs_week | toy_type), data = toys_read)
anova(restricted_model, full_model, test = "Chisq")
```

:::int
__Reporting__  

A likelihood ratio test indicated that the inclusion of weekly hours of reading practice as a predictor was associated with an improvement in model fit ($\chi^2(1) = 11.81, p < .001$).
:::

`r optend()`
`r optbegin("Optional: Parametric Bootstrap LRT",olabel=F,toggle=params$TOGGLE)`  
There are also various "bootstrapping" methods which it is worth looking into. Think back to USMR when we first learned about hypothesis testing. Remember that we did lots of simulating data, so that we can compare what we actually observe with what we would expect if the null hypothesis were true? By doing this, we were essentially _creating_ a null distribution, so that our calculating a p-value can become an issue of summarising data (e.g. calculate the proportion of our simulated null distribution that is more extreme than our observed statistic)

Instead of assuming that the likelihood ratio test statistics are $\chi^2$-distributed, we can bootstrap this test instead. This approach simulates data from the simpler model, fits both the simple model and the complex model and evaluates the change in log-likelihood. By doing this over and over again, we build a distribution of what changes in log-likelihood we would be likely to see if the more complex model is not any better. In this way it actually constructs a distribution reflecting our null hypothesis, against which we can then compare our actual observed effect
```{r echo=F}
set.seed(993)
```
```{r eval=FALSE}
library(pbkrtest)
PBmodcomp(full_model, restricted_model, nsim=1000)
```

:::int
__Reporting__  

A parametric bootstrap likelihood ratio test (R = 1000) indicated that the inclusion of weekly hours of reading practice as a predictor was associated with an improvement in model fit ($LRT = 11.79, p = .004$).
:::

`r optend()`
`r optbegin("Optional: Parametric Bootstrap Confidence Intervals",olabel=F,toggle=params$TOGGLE)`  
Much the same as above, but with just one model we simulate data many times and refit the model, so that we get an empirical distribution that we can use to construct confidence intervals for our effects. 
```{r eval=FALSE}
confint(full_model, method="boot")  
```

:::int
__Reporting__  

95% Confidence Intervals were obtained via parametric bootstrapping with 1000 iterations.  
...  
...   
Weekly hours of reading practice was associated increased reading age ($\beta = 1.14,\ 95%\ CI\ [0.58 -- 1.73]$).

:::

`r optend()` 
`r optbegin("Optional: Case-based Bootstrap Confidence Intervals",olabel=F,toggle=params$TOGGLE)`
It's worth noting that there are many different types of bootstrapping that we can conduct. Different methods of bootstrapping vary with respect to the assumptions we will have to make when using them for drawing inferences. For instance, the parametric bootstrap discussed above assumes that explanatory variables are fixed and that model specification and the distributions such as $\zeta_i \sim N(0,\sigma_{\zeta})$ and $\varepsilon_i \sim N(0,\sigma_{\varepsilon})$ are correct.  
An alternative is to generate a distribution by __resampling with replacement__ from our data, fitting our model to the resample, and then repeating this over and over. This doesn't have to rely on assumptions about the shape of the distributions of $\zeta_i$ and $\varepsilon_i$ - we just need to ensure that we correctly specify the hierarchical dependency of data. It does, however, require the decision of at which levels to resample (this is discussed more in week 4, and is what the __lmeresampler__ package is all about).    
`r optend()`

:::frame 

If you want more information (not required reading for this course), then Julian Faraway has a page [here](https://people.bath.ac.uk/jjf23/mixchange/index.html) with links to some worked examples, and Ben Bolker has a wealth of information on his [GLMM FAQ pages](http://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#why-doesnt-lme4-display-denominator-degrees-of-freedomp-values-what-other-options-do-i-have).  

:::

<!-- Formatting -->


<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
