---
title: "Multilevel Models for Longitudinal Data"
params: 
    SHOW_SOLS: FALSE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(xaringanExtra)
library(tidyverse)
library(patchwork)
xaringanExtra::use_panelset()
```


multilevel models are perfectly geared towards dealing with longitudinal data, which can be framed as having essentially the same hierarchical structure as "children in schools".  




::: {.callout-note collapse="true"}
#### Mindful Cognitive Aging

A study is interested in examining **whether engaging in mindfulness can prevent cognitive decline in older adults**. They recruit a sample of 20 participants at age 60, and administer the Addenbrooke's Cognitive Examination (ACE) every 2 years (until participants were aged 78). Half of the participants undertake to complete daily mindfulness sessions, while the remaining participants did not.   

**DATASET:** [https://uoepsy.github.io/data/dapr3_mindfuldecline.csv](https://uoepsy.github.io/data/dapr3_mindfuldecline.csv)

```{r}
#| echo: false
d3 <- read_csv("https://uoepsy.github.io/data/dapr3_mindfuldecline.csv")
tibble(
  variable=names(d3),
  description=c("Site Identifier",
                "Participant Identifier",
                "Whether the participant engages in mindfulness or not (control/mindfulness)",
                "Visit number (1 - 10)",
                "Age (years) at visit",
                "Addenbrooke's Cognitive Examination Score. Scores can range from 0 to 100",
                "Clinical diagnosis of cognitive impairment ('imp' = impaired, 'unimp' = unimpaired)")
) |> gt::gt()

```

:::