---
title: "Centering regression predictors"
author: "jk"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
pkgdown:
  as_is: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
set.seed(987)
tibble(
  id = letters[1:10],
  start = seq(5,20,length.out=10),
  end = seq(15,30,length.out=10),
  coef = rnorm(10,-2)
) %>% mutate(
  x = map2(start,end, ~runif(10,.x,.y)),
  y = map2(x, coef,~.x*.y)
) %>% unnest %>%
  mutate(y=y+10*start+rnorm(n(),0,3)) -> df

# ggplot(df,aes(x=x,y=y,col=id))+geom_point(
```



# centering in LM

```{r echo=FALSE}
mod = lm(y~x,df)
p1 = ggplot(df, aes(x=x,y=y))+geom_point()+
  geom_smooth(method="lm",se=F)+
  geom_smooth(method="lm",se=F,fullrange=T,lty="dotted")+
  xlim(0,30)+ylim(0,max(df$y))+
  geom_segment(x=0,xend=0,y=0,yend=max(df$y))+
  geom_segment(x=0,xend=30,y=0,yend=0)+
  geom_point(data=tibble(x=0,y=coef(mod)[1]),size=4)

  
  
mod = lm(y~scale(x),df)
p2 = ggplot(df, aes(x=x,y=y))+geom_point()+
  geom_smooth(method="lm",se=F)+
  geom_smooth(method="lm",se=F,fullrange=T,lty="dotted")+
  ylim(0,max(df$y))+
  geom_segment(x=mean(df$x),xend=mean(df$x),y=0,yend=max(df$y))+
  geom_segment(x=0,xend=30,y=0,yend=0)+
  geom_point(data=tibble(x=mean(df$x),y=coef(mod)[1]),size=4)+
  scale_x_continuous(limits=c(0,30),
                     breaks=c(mean(df$x)-(2*sd(df$x)), mean(df$x)-sd(df$x), 
                              mean(df$x), 
                              mean(df$x)+sd(df$x), mean(df$x)+(2*sd(df$x))),
                     labels=c(-2,-1,0,1,2))

p1 + p2
```

# Centering predictors in MLM

```{r}
ggplot(df,aes(x=x,y=y))+
  geom_smooth(method="lm",se=F)+
  geom_point(aes(col=id))+
  geom_smooth(method="lm",se=F, aes(col=id))
  
```

- grand mean centering in MLM
- group mean centering in MLM "for a .."
- adding in group means as an L2 predictor
- what is a "context effect" ludtke?
- latent variable mlms? croon


Lüdtke, O., Marsh, H. W., Robitzsch, A., Trautwein, U., Asparouhov, T., & Muthén, B. (2008). The multilevel latent covariate model: A new, more reliable approach to group-level effects in contextual studies. Psychological Methods, 13, 203–229.

Croon, M. A., & van Veldhoven, M. J. P. M. (2007). Predicting group-level outcome variables from variables measured at the individual level: A latent variable multilevel model. Psychological Methods, 12, 45–57.




